Mermory-efficient Fujisaki-Okamoto transformation

### 2024-05-08 lattices/compact-fo-using-mac/manuscript/main.tex
checkpoint

% Common shortcuts
\newcommand{\round}[1]{\lfloor {#1} \rceil}
\newcommand{\Norm}[1]{\Vert {#1} \Vert}
\newcommand{\norm}[1]{\vert {#1} \vert}
\newcommand{\var}[1]{\operatorname{Var}[{#1}]}
\newcommand{\leftsample}{\overset{{\scriptscriptstyle\$}}{\leftarrow}}
\newcommand{\keygen}{\operatorname{KeyGen}}

@@

\section{OW-PCVA transformation}
Let $(\keygen, E, D)$ define a probabilistic public-key encryption scheme, then $T$ is a transformation that takes this scheme and outputs a second public-key encryption scheme. The transformation makes use of two additional components:

\begin{enumerate}
    \item A $\operatorname{MAC}: \mathcal{K}^\text{MAC} \times \{0,1\}^\ast \rightarrow \mathcal{T}$
    \item A hash function $G: \mathcal{M} \rightarrow \text{Coin} \times \mathcal{K}^\text{MAC}$
\end{enumerate}

\begin{multicols}{2}
    \begin{algorithm}[H]
        \SetAlgoLined
        \caption{OW-PCVA encryption $E^T$}
        \KwIn{
            $\text{pk}$, $m \in \mathcal{M}$
        }
        $(r, k) \leftarrow G(m)$\;
        $c \leftarrow E(\text{pk}, m, r)$\;
        $t \leftarrow \operatorname{MAC}(k, c)$
        \tcp*[h]{"encrypt-then-mac"}\;
        \Return{$(c, t)$}\;
    \end{algorithm}

@@

    \begin{algorithm}[H]
        \SetAlgoLined
        \caption{OW-PCVA decryption $D^T$}
        \KwIn{
            $\text{sk}, c \in \mathcal{C}, t \in \mathcal{T}$
        }
        $\hat{m} \leftarrow D(\text{sk}, c)$\;
        $(\hat{r}, \hat{k}) \leftarrow G(\hat{m})$\;
        \If{
            $\operatorname{MAC}(\hat{k}, c) \neq t$
        }{
            \Return{$\bot$}\;
        }
        \Return{$\hat{m}$}\;
    \end{algorithm}
\end{multicols}

Instead of using re-encryption to check for the integrity of the ciphertext, a MAC tag is used. At the cost of adding a fixed number of bytes to the ciphertext, the decryption routine becomes significantly faster and memory-efficient. The memory trade-off is especially important since in Kyber, re-encryption needs to compute the entire public key $A \in R_q^{k \times k}$ from the 32-byte seed.

We claim that this transformation is still OW-PCVA. The proof will be largely similar to that of the original OW-PCVA transformation proof, but with some notable differences:

\begin{itemize}
    \item When queried on some authenticated ciphertext $(\tilde{c}, \tilde{t})$, the secret-key-less $\operatorname{CVO}_1$ searches the hash oracle $G$ for $(m, r, k)$ such that $\operatorname{MAC}(k, \tilde{c}) = \tilde{t}$. The diverging event between using $\operatorname{CVO}$ and using $\operatorname{CVO}_1$ is ``adversary produces valid authenticated ciphertext without consulting the hash oracle'', which is equivalent to selective forgery (or existential forgery, I am not sure which): $\epsilon_a - \epsilon_b \leq q_V \cdot \epsilon_\text{MAC}$
\end{itemize}

\begin{theorem}\label{mac-fo-theorem}
    Let $(\keygen, E, D)$ be a public-key encryption scheme that is $\delta$-correct and that has $\gamma$-spread. For every OW-PCVA adversary against the transformed encryption scheme $(E^T, D^T)$ that makes $q_G$ hash queries and $q_V$ ciphertext validation queries and that has advantage $\epsilon_\text{OW-PCVA}$, there exists an IND-CPA adversary against $(\keygen, E, D)$ with advantage $\epsilon_\text{IND-CPA}$ and an EF-CMA adversary against the MAC with advantage $\epsilon_\text{MAC}$ such that:

    \begin{equation*}
        \epsilon_\text{OW-PCVA}
        \leq q_G \cdot \delta 
        + q_V \cdot \epsilon_\text{MAC}
        + \frac{2q+1}{\norm{\mathcal{M}}} 
        + \epsilon_\text{IND-CPA}
    \end{equation*}
\end{theorem}

The proof will be discussed in section \ref{proof-of-mac-fo-theorem}.

\section{Proof techniques}
Similar to the 2017 paper, we will use a sequence of games to incrementally replace PCO, CVO, and other aspects of the standard OW-PCVA game with some simulation. Then we will show that the total loss of security is negligible.

\subsection{Replacing plaintext-checking oracle}\label{replacing-pco}
The plaintext-checking oracle $\operatorname{PCO}$ takes as input a plaintext-ciphertext pair $(m, (c, t))$ and return ``reject'' if and only if the the ciphertext is a valid encryption of the plaintext and vice versa under the context of some fixed keypair.

The simulated plaintext-checking oracle $\operatorname{PCO}_1$ removes the requirement for the secret key by removing the step that checks whether the queried ciphertext $(c, t)$ decrypts back to the queried plaintext $m$.

From the OW-PCVA adversary's perspective, the two oracles are indistinguishable, except for when there is a decryption error $D(\text{sk}, E(\text{pk}, m, r)) \neq m$ for the input plaintext $m$ and the corresponding coin $r$. The MAC is not involved in this argument. The loss of security is still $q_G\cdot\delta$

\begin{multicols}{2}
    \begin{algorithm}[H]
        \SetAlgoLined
        \caption{Vanilla PCO}
        \KwIn{
            $m \in \mathcal{M}, c \in \mathcal{C}, t \in \mathcal{T}$ 
        }
        \If{
            $D^T(\text{sk}, (c, t)) \neq m$
        }{
            \Return{$\bot$}\;
            \tcp*[h]{Decryption did not match}
        }
        \If{
            $E^T(\text{pk}, m) \neq (c, t)$
        }{
            \Return{$\bot$}\;
            \tcp*[h]{Encryption did not match}
        }
    \end{algorithm}

    \begin{algorithm}[H]
        \SetAlgoLined
        \caption{Simulated $\operatorname{PCO}_1$}
        \KwIn{
            $m \in \mathcal{M}, c \in \mathcal{C}, t \in \mathcal{T}$
        }
        \If{
            $E^T(\text{pk}, m) \neq (c, t)$
        }{
            \Return{$\bot$}\;
            \tcp*[h]{Encryption did not match}
        }
    \end{algorithm}
\end{multicols}

\subsection{Replacing ciphertext-validation oracle}\label{replacing-cvo}
This is a similar trick to what was used in the 2017 paper, as well. Where the ciphertext is honestly generated, both oracles will return accept. Where the vanilla CVO rejects the ciphertext (aka ciphertext is malformed or the tag doesn't match), the simulated CVO will also reject. Therefore, the diverging event is when the vanilla CVO accepts but the simulated CVO rejects. Since the vanilla CVO accepts the query, the tag $t$ must be valid for the queried ciphertext $c$; on the other hand, the simulated CVO's rejection means there is no matching query in the hash oracle. Under the random oracle assumption, $t$ is a valid tag for some data $c$ under some key that the adversary does not know. In other words, $(c, t)$ is some kind of forgery.

\textbf{The argument on how the security of the MAC relates to this diverging event is still a bit fuzzy}, but here are two possible ways I can think of:

\begin{itemize}
    \item For each CVO query, the adversary is trying to forge tag for that specific ciphertext. This means there is a selective forgery attack, and over all $q_V$ validation queries, the probability of having at least one selective forgery attack that works is at most $q_V\epsilon_\text{SF}$

    \item Across all CVO query, the adversary wants to forge tag for some ciphertext, meaning there is a existential forgery attack, and the probability is at most $\epsilon_\text{EF}$
\end{itemize}

\begin{multicols}{2}
    \begin{algorithm}[H]
        \SetAlgoLined
        \caption{Vanilla CVO}
        \KwIn{
            $(c \in \mathcal{C}, t \in \mathcal{T})$
        }
        \If{
            $D^T(\text{sk}, (c, t)) = \bot$
        }{
            \Return{$\bot$}\;
        }
    \end{algorithm}

    \begin{algorithm}[H]
        \SetAlgoLined
        \caption{Simulated $\operatorname{CVO}_1$}
        \KwIn{
            $(c \in \mathcal{C}, t \in \mathcal{T})$
        }
        \If{
            $\exists (m, r, k) \in \mathcal{O}^G$ such that $\operatorname{MAC}(k, c) = t$
        }{
            \Return{Accept}\;
        }
        \Return{$\bot$}
    \end{algorithm}
\end{multicols}

\subsection{Random until queried}\label{random-until-queried}
We then replace the pseudorandom coin and the MAC key with truly random coins and truly random keys in the challenge encryption. These two challenge encryption routines are identical, unless the adversary queries $G$ with the challenge plaintext $m^\ast$, but the probability of making such query can be bounded.

\begin{multicols}{2}
    \begin{algorithm}[H]
        \SetAlgoLined
        \caption{Challenge encryption}
        $m^\ast \leftsample \mathcal{M}$\;
        $(c^\ast, t^\ast) \leftarrow E^T(\text{pk}, m^\ast)$\;
        \Return{$(c^\ast, t^\ast)$}
    \end{algorithm}

    \begin{algorithm}[H]
        \SetAlgoLined
        \caption{Simulated challenge encryption}
        $m^\ast \leftsample \mathcal{M}$\;
        $r^\ast \leftsample \text{Coin}, k^\ast \leftsample \mathcal{K}_\text{MAC}$\;
        $c^\ast \leftarrow E(\text{pk}, m^\ast, r^\ast)$\;
        $t^\ast \leftarrow \operatorname{MAC}(k^\ast, c^\ast)$
        \Return{$(c^\ast, t^\ast)$}
    \end{algorithm}
\end{multicols}

Let the OW-PCVA adversary's advantage using the vanilla challenge encryption be $\epsilon_0$ and its advantage using the simulated challenge encryption be $\epsilon_1$. Denote the event that the adversary queries the hash oracle $G$ with the challenge plaintext $m^\ast$ by $\operatorname{QUERY}^\ast$, then:

\begin{equation*}
    \epsilon_0 - \epsilon_1 \leq P[\operatorname{QUERY}^\ast]
\end{equation*}

We can build an IND-CPA adversary against the underlying encryption scheme to bound $P[\operatorname{QUERY}^\ast]$: if the OW-PCVA adversary makes the magic query, then the IND-CPA adversary can find it in the hash oracle's tape and win the IND-CPA game; if the OW-PCVA adversary does not make the magic query, then the IND-CPA adversary outputs a blind guess:

\begin{equation*}
    P[\operatorname{QUERY}^\ast] \leq \epsilon_\text{IND-CPA}
\end{equation*}

\section{Complete proof of \ref{mac-fo-theorem}}\label{proof-of-mac-fo-theorem}
\begin{proof}
    We will prove using a sequence of games. Game 0 is the standard OW-PCVA game.

    Game 1 is identical to game 0, except PCO is replaced with $\operatorname{PCO}_1$, the loss of security is described in section \ref{replacing-pco}:

    \begin{equation*}
        \epsilon_0 - \epsilon_1 \leq q_G \cdot \delta
    \end{equation*}

    Game 2 is identical to game 1, except CVO is replaced with $\operatorname{CVO}_1$, the loss of security is described in section \ref{replacing-cvo}. Here I put a placeholder $\epsilon_\text{MAC}$ to indicate that this quantity is tied to the security of the MAC, but I don't have a solid answer yet

    \begin{equation*}
        \epsilon_1 - \epsilon_2 \leq \epsilon_\text{MAC}
    \end{equation*}

    Game 3 is identical to game 2, except the challenge encryption is replaced with the simulated challenge encryption

    \begin{equation*}
        \epsilon_3 - \epsilon_2 \leq P[\operatorname{QUERY}^\ast]
    \end{equation*}

    From section \ref{random-until-queried} we know that $P[\operatorname{QUERY}^\ast] \leq \epsilon_\text{IND-CPA}$ for some IND-CPA adversary against the underlying PKE.

    Game 3 can be entirely simulated by an OW-CPA adversary against the underlying PKE:

    \begin{itemize}
        \item The keypairs are identical between the two PKE's
        \item The OW-CPA adversary can simulate PCO, CVO, and hash oracle $G$
        \item The OW-CPA adversary receives challenge ciphertext $c^\ast$ which is computed from a truly random coin; it can then random a truly random MAC key and compute the tag $t^\ast$
        \item The OW-CPA adversary passes $(c^\ast, t^\ast)$ to the OW-PCVA adversary and returns whatever the OW-PCVA adversary returns
    \end{itemize}

    The OW-CPA adversary wins if and only if the OW-PCVA adversary wins: $\epsilon_3 = \epsilon_\text{OW-CPA}$.

    Putting everything together we have:

    \begin{equation*}
        \epsilon_0 
        \leq q_G \cdot \delta 
        + \epsilon_\text{MAC} 
        + \epsilon_\text{IND-CPA} 
        + \epsilon_\text{OW-CPA}
    \end{equation*}
\end{proof}

>>>>

% Common shortcuts
\newcommand{\round}[1]{\lfloor {#1} \rceil}
\newcommand{\Norm}[1]{\Vert {#1} \Vert}
\newcommand{\norm}[1]{\vert {#1} \vert}
\newcommand{\var}[1]{\operatorname{Var}[{#1}]}
\newcommand{\leftsample}{\overset{{\scriptscriptstyle\$}}{\leftarrow}}
% Cryptographic primitives
\newcommand{\pke}{\operatorname{PKE}}
\newcommand{\enc}{\operatorname{Enc}}
\newcommand{\dec}{\operatorname{Dec}}
\newcommand{\mac}{\operatorname{MAC}}
\newcommand{\cvo}{\operatorname{CVO}}
\newcommand{\pco}{\operatorname{PCO}}
\newcommand{\keygen}{\operatorname{KeyGen}}
\newcommand{\llbrack}{[\![}
\newcommand{\rrbrack}{]\!]}

@@

\section{Introduction}

\section{Preliminaries}

\subsection{Security definitions}

\subsection{Fujisaki-Okamoto transformation}

\section{Memory-efficient OW-PCVA transformation using MAC}
In this section we propose an alternative OW-PCVA transformation $\operatorname{T}$.

Let $\pke = (\keygen, \enc, \dec)$ be a public-key encryption scheme defined over key space $\mathcal{K}_{\pke}$, message space $\mathcal{M}$, ciphertext space $\mathcal{C}$, and coin space $\mathcal{R}$. Let $\mac = (\operatorname{Sign}, \operatorname{Verify})$ be a message authentication code defined over key space $\mathcal{K}_{\mac}$, message space $\mathcal{M}$ (the same message space as $\pke$), and tag space $\mathcal{T}$. Let $G: \mathcal{M} \rightarrow \mathcal{R} \times \mathcal{K}_{\mac}$ be a random oracle. The transformed public-key encryption scheme is denoted by $\pke_1 = T(\pke, \mac, G)$ and includes the sub-routines $(\keygen, \enc_1, \dec_1)$.

The transformed encryption and decryption routines are as follows

\begin{multicols}{2}
    \begin{algorithm}[H]
        \SetAlgoLined
        \caption{OW-PCVA encryption $E_1$}
        \KwIn{
            $\text{pk}$, $m \in \mathcal{M}$
        }
        $(r, k) \leftarrow G(m)$\;
        $c \leftarrow E(\text{pk}, m, r)$\;
        $t \leftarrow \operatorname{MAC}(k, c)$
        \tcp*[h]{"encrypt-then-mac"}\;
        \Return{$(c, t)$}\;
    \end{algorithm}

@@

    \begin{algorithm}[H]
        \SetAlgoLined
        \caption{OW-PCVA decryption $D_1$}
        \KwIn{
            $\text{sk}, c \in \mathcal{C}, t \in \mathcal{T}$
        }
        $\hat{m} \leftarrow D(\text{sk}, c)$\;
        $(\hat{r}, \hat{k}) \leftarrow G(\hat{m})$\;
        \If{
            $\operatorname{MAC}(\hat{k}, c) \neq t$
        }{
            \Return{$\bot$}\;
        }
        \Return{$\hat{m}$}\;
    \end{algorithm}
\end{multicols}

If $\pke$ is OW-CPA secure and $\mac$ is existentially unforgeable under chosen message attack, then the transformed encryption scheme $\pke_q$ is OW-PCVA:

\begin{theorem}
    Let $\pke$ be $\delta$-correct and have $\gamma$-spread. For every OW-PCVA adversary $\mathcal{A}_1$ against $\pke_1$ that makes $q_V$ ciphertext validation queries and $q_G$ hash queries, and that has advantage $\epsilon_1$, there exists an OW-CPA adversary against $\pke$ with advantage $\epsilon_{\pke}$ and an EF-CMA adversary against $\mac$ with advantage $\epsilon_{\mac}$ such that

    \begin{equation*}
        \epsilon_{\pke_1} \leq q_G \cdot \delta + q_V \cdot \epsilon_{\mac} + q_G \cdot \epsilon_{\pke}
    \end{equation*}
\end{theorem}

The proof follows a sequence of games that begins with the standard OW-PCVA game but incrementally replaces the PCO and the CVO with alternative implementations that can be simulated by an OW-CPA adversary and/or the EF-CMA adversary who does not have the secret key.

\begin{proof}
    Let $\mathcal{A}$ denote an adversary against the OW-PCVA security of $\enc_1$. Consider the following sequence of games.

    Game $G_0$ is the standard OW-PCVA game. In the original OW-PCVA game, queries to the random oracle $G$ are stored on a tape $\mathcal{L}^G$. By convention, we say that $G(m) = (r, k)$ if and only if $(m, r, k) \in \mathcal{L}^G$.

    We can be more explicit with the implementation of the PCO and the CVO:

    \begin{multicols}{2}
        \begin{algorithm}[H]
            \SetAlgoLined
            \caption{
                Standard $\pco$ in $G_0$
            }
            \KwIn{
                $m \in \mathcal{M}, (c \in \mathcal{C}, t \in \mathcal{T})$
            }
            $\hat{m} \leftarrow \dec(\operatorname{sk}, c)$\;
            $(\hat{r}, \hat{k}) = G(\hat{m})$\;

            \Return{
                $\llbrack m = \hat{m} \rrbrack$ 
                and $\llbrack \mac(\hat{k}, c) = t\rrbrack$
            }
        \end{algorithm}

        \begin{algorithm}[H]
            \SetAlgoLined
            \caption{
                Standard $\cvo$ in $G_0$
            }
            \KwIn{
                $c \in \mathcal{C}, t \in \mathcal{T}$
            }
            $\hat{m} \leftarrow \dec(\operatorname{sk}, c)$\;
            $(\hat{r}, \hat{k}) \leftarrow G(\hat{m})$\;
            \Return{$\llbrack \mac(\hat{k}, c) = t\rrbrack$}
        \end{algorithm}
    \end{multicols}

    Game $G_1$ is identical to $G_0$, except that the CVO is replaced with an alternative implementation. Instead of executing the decryption routine $\hat{m} \leftarrow \dec(\operatorname{sk}, (c, t))$ to recover the plaintext, the alternative implementation checks previous hash queries in $\mathcal{L}^G$ to recover the plaintext.

    \begin{algorithm}[H]
        \SetAlgoLined
        \caption{
            $\cvo_1$ in $G_1$
        }
        \KwIn{
            $(c \in \mathcal{C}, t \in \mathcal{T})$
        }
        \If{
            $\exists (m, r, k) \in \mathcal{L}^G$ such that
            $\mac(k, c) = t$
        }{
            \Return{$1$}\;
        }
        \Return{$\bot$}\;
    \end{algorithm}

    Consider a single ciphertext query $(c, t)$. If the query is accepted by $\cvo_1$, then this cipehrtext query is honestly generated, so the correctness of $\pke$ and $\mac$ implies that the query will also be accepted by $\cvo$. Similarly, if the query is rejected by $\cvo$, then either the re-encryption does not match, or the MAC does not match; in any case, $\cvo_1$ will also reject the ciphertext query. Therefore, these two implementations disagree if and only if $\cvo$ accepts the query but $\cvo_1$ rejects the query.

    Since $\cvo$ accepts the query, $t$ is a valid tag for $c$ under some MAC key $k$ derived by hashing the decryption $m \leftarrow \dec(\operatorname{sk}, c)$. Since there is no matching query in $\mathcal{L}^G$, under the random oracle assumption, the MAC key $k$ is indistinguishable from truly random. In other words, for some ciphertext $c$ of the adversary's choosing, $t$ is a forgery under an unknown key. Therefore, for a single ciphertext query, the probability that $\cvo$ and $\cvo_1$ disagree is bounded by the advantage of some existential forgery adversary: $P[\cvo(c, t) \neq \cvo_1(c, t)] \leq \epsilon_{\mac}$.

    Game 1 and game 0 behave differently if one or more among all $q_V$ ciphertext validation queries satisfies the condition above, therefore:

    \begin{equation*}
        \epsilon_0 - \epsilon_1 \leq q_V \cdot \epsilon_{\mac}
    \end{equation*}

    Game 2 $G_2$ is identical to $G_1$, except that the standard $\pco$ is replaced by an alternative implementation $\pco_1$

    \begin{algorithm}[H]
        \SetAlgoLined
        \caption{$\pco_1$ in $G_2$}
        \KwIn{
            $m \in \mathcal{M}, (c \in \mathcal{C}, t \in \mathcal{T})$
        }
        $(r, k) \leftarrow G(m)$\;
        \Return{
            $\llbrack \enc(\operatorname{pk}, m; r) = c \rrbrack$
            and $\llbrack \mac(k, c) = t \rrbrack$
        }
    \end{algorithm}

    % TODO: 
\end{proof}

\section{Experimental results}
